---
title: "Simulation study"
author: "Michael A. Spence"
date: "30/07/2020"
output:
  pdf_document: default
bibliography: bibfile.bib
---
  
<style>
body {
text-align: justify;
}
</style>
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
```

```{r, echo = FALSE, message = FALSE}
# Install and load required packages
req_packages <- c("LeMaRns", "randtoolbox", "parallel")
for (i in req_packages) {
  if (!i %in% installed.packages()[, "Package"]) {install.packages(i)}
  suppressPackageStartupMessages(suppressWarnings(require(i, character.only = TRUE)))
}
```

# Introduction

In this document I demonstrate a simulation study that uses the methods described in the manuscript to fit a simulated example using the LeMans model [@thorpe15] with 3 species. For a complete description of the model see @LeMaRns. I ran the model for 20 years and observed the commerial catches, $w_{i,t}$, and survey catches, $z_{i,t}$, for each species $i=1,2,3$ and year $t=1,\ldots{},20$. The uncertain static parameters are $b_i$ (or `b`) from the `hockey stick` recruitment curve; $\phi_{0,i}$ (or `spinupF`), the fishing mortality during the spin up period and $\kappa$ (or `other`), the other food available in the system. In addition I am interested in the dynamical parameters, the fishing rates for the 20 species each year, $\phi_{t,i}$, for $t=1\ldots{}20$ and $i=1,2,3$.

The likelihood of the model is
\begin{equation*}
\prod{}_{t=1}^{20}\prod{}_{i=1}^{3}N(\ln(w_{i,t})|\ln(c(M_{i,t}(\theta,\phi_{1:20,1:3}))), \sigma_{c,i}^2)N(\ln(z_{i,t})|\ln(b(M_{i,t}(\theta,\phi_{1:20,1:3}))), \sigma_{s,i}^2),
\end{equation*}
where $c(M_{i,t}(\theta,,\phi_{1:20,1:3}))$ is the commercial catch of the $i$-th species in the $t$-th year, $b(M_{i,t}(\theta,,\phi_{1:20,1:3}))$ is the survey catch of the $i$-th species in the $t$-th year, $\theta=(\ln(\kappa),\ln(b_1),\ln(b_2),\ln(b_3),\phi_{0,1},\phi_{0,2},\phi_{0,3})$ and $N(a|d,e)$ is a normal density with expectation $d$ and variance $e$ evaluated at $a$. $\sigma_{c,i}^2$ and $\sigma_{s,i}^2$ will be estimated.

I believed that $\ln(\kappa)\in(0,30)$, $\ln(b_i)\in(0,\infty)$ and $\phi_{t,i}$ for $i=1,2,3$ and $t=0,1,\ldots{}20$. I also believed
\begin{equation*}
\sigma_{c,i}^2\sim{}\text{Inv-Gamma}(0.1,0.1)
\end{equation*}
and 
\begin{equation*}
\sigma_{s,i}^2\sim{}\text{Inv-Gamma}(0.1,0.1)
\end{equation*}
for $i=1,2,3$.


# Simulating the data
First I load the package and setup a model with only three species in it, sprat, mackerel and cod:
```{r,eval=T}
library(LeMaRns)
set.seed(14)
spec_par <- NS_par[c(1,11,20),]
tau <- NS_tau[c(1,11,20),c(1,11,20)]
spec_par$a <- exp(17.429 - 3.792 * log(spec_par$Linf))
```
The set of the parameters that I am going to try and fit.
```{r,eval=T}
spec_par$b <- exp(c(11,7.5,1))
```
therefore the true value of $\theta_{2:4}$ will be `c(11,7.5,1)`. I then setup the model with $\theta_1$ equal to 27:
```{r,eval=T,warning=FALSE}
params <- LeMansParam(spec_par, tau=tau, eta=rep(0.25, 3), L50=spec_par$Lmat, other=exp(27))
```
and the effort matrix for the spin-up period, which in this case is 100 years:
```{r,eval=T}
effort <- matrix(c(0.1,0.3,0.6), 100, dim(params@Qs)[3],byrow = T)
```
The fishing effort during the spin up period, $\theta_{5:7}$ is `c(0.1,0.3,0.6)` for each of the three species, a parameter I am interested in fitting. I simulate the spin up period:
```{r,eval=T}
model_run <- run_LeMans(params, years=100, effort=effort)
```

For the fishing period I need to add the suvey vessel. I use data from @walker_selection and setup the gear selection:
```{r,eval=T}
custom_q <- matrix(0,32,3)
custom_q[,1] <- c(0.017254858,0.043039643,0.284392929,rep(0.430695848,29))
custom_q[,2] <- c(0.096113412,rep(0.24028353,31))
custom_q[,3] <- c(0.001761793,0.006166274,0.039181631,0.092776614,0.145237111,0.193340804,
                  0.232928633,0.2564465,0.27551981,0.28422822,0.290723986,0.296690919,
                  0.305572708,0.316083176,0.334233041,0.358307207,0.388331838,0.416365328,
                  0.455782759,0.499352575,0.536897304,0.587032109,0.640318341,0.695650091,
                  0.740381939,0.740381939,0.740381939,0.740381939,0.740381939,0.740381939,
                  0.740381939,0.740381939
)
params@Qs <- abind::abind(params@Qs,custom_q)
dimnames(params@Qs)[[3]][4] <- "survey"
```
I am going to run a 20 year time series with fishing effort and survey effort:
```{r,eval=T}
Fs <- abs(cbind(abs(cumsum(rnorm(20,0,0.075)) + 0.1),cumsum(rnorm(20,0,0.075)) 
                + 0.6,cumsum(rnorm(20,0,0.075)) + 0.25))
sur_ef <- abs(rnorm(20,0.0001,0.0001))
fish_rate <- as.matrix(cbind(Fs,sur_ef));colnames(fish_rate) <- NULL
```
The fishing effort is
```{r,eval=T}
Fs
```
I then simulate the catch data:
```{r,eval=T}
model_run_act <- run_LeMans(params, years=20, effort=fish_rate,N0=model_run@N[,,1001])
plot_biomass(params,model_run_act)
catches <- get_CPG(params,model_run_act,effort=fish_rate)
com_catches_t <- apply(catches[,1:3,],c(2,3),sum)
sur_catch <-  com_catches <- matrix(0,20,3)
for (i in 1:20){
  com_catches[i,] <- log(rowSums(com_catches_t[,10 * (i-1) + (1:10)])) - log(1e6)
  sur_catch[i,] <- log(rowSums(catches[,4,10 * (i-1) + (1:10)])) - log(1e6)
}
```
and add some stochasticity to the catches
```{r,eval=T}
sigma_c <- c(0.38,0.44,0.5); sigma_s <- c(0.6,0.7,0.6)
obs_c <- com_catches + matrix(rnorm(60,0,sigma_c),20,3,byrow=T)
obs_s <- sur_catch + matrix(rnorm(60,0,sigma_s),20,3,byrow=T)
```
The observations are:
```{r,eval=T}
par(mfrow=c(3,2))
plot(1:20,obs_c[,1],main="Sprat",ylab="log commercial catches",type="l")
plot(1:20,obs_s[,1],main="Sprat",ylab="log survey catches",type="l")
plot(1:20,obs_c[,2],main="Mackerel",ylab="log commercial catches",type="l")
plot(1:20,obs_s[,2],main="Mackerel",ylab="log survey catches",type="l")
plot(1:20,obs_c[,3],main="Cod",ylab="log commercial catches",type="l")
plot(1:20,obs_s[,3],main="Cod",ylab="log survey catches",type="l")
```

# History matching

To do the history matching to find initial values for the MCMC algorithm I first need to setup two functions. The first takes a model output and calculates the catches:
```{r,eval=F}
calc_catch <- function(params,model_run,effort){
  catches <- get_CPG(params,model_run,effort=effort)
  com_catches_t <- apply(catches[,1:3,],c(2,3),sum)
  times <- (dim(model_run@N)[3] - 1)*params@phi_min
  sur_catch <-  com_catches <- matrix(0,times,3)
  for (i in 1:times){
    com_catches[i,] <- log(rowSums(com_catches_t[,10 * (i-1) + (1:10)])) - log(1e6)
    sur_catch[i,] <- log(rowSums(catches[,4,10 * (i-1) + (1:10)])) - log(1e6)
  }
  return(list(com=com_catches,sur=sur_catch))
}
```
and the second takes in the uncertain static parameters, $\theta$ and the fishing rates of the first year and outputs the simulated catches from the first year, $F_{1,i}$:
```{r,eval=F}
hist_mat <- function(input,params){
  params@other <- exp(input[1])
  params@recruit_params[[1]][2] <- exp(input[2])
  params@recruit_params[[2]][2] <- exp(input[3])
  params@recruit_params[[3]][2] <- exp(input[4])
  model_run <- run_LeMans(params, years=100, effort=t(matrix(c(input[5:7],0),4,100)))
  model_run1 <- run_LeMans(params,years=1,effort=t(matrix(c(input[8:10],sur_ef[1]),4,1)),
                           N0=model_run@N[,,dim(model_run@N)[3]])
  effort <- t(matrix(c(input[8:10],sur_ef[1]),4,1))
  return(unlist(calc_catch(params,model_run1,effort)))
}
```
I now need to generate parameter values for the first round of the history matching using Sobol' indices @sobol and run them:
```{r,eval=F}
library(randtoolbox) ## for sobol indicies
sam1 <- sobol(5000,3*3 + 1)
## other - between 20 and 30
sam1[,1] <- sam1[,1] *20 +10
## b between 0 and 20
sam1[,2:4] <- sam1[,2:4] * 20
## spin-up F and first year fishing mortality betweem 0 and 1.5
sam1[,5:10] <- sam1[,5:10] * 1.5
library(pbapply) ## do in parallel
round1 <- pbapply(sam1,1,hist_mat,params=params)
```
```{r,echo=F}
load("after_2_rounds.Rdata")
```
I then compare the difference between the observed values and the simulated values
```{r,eval=T}
c_r1 <- round1 - c(obs_c[1,],obs_s[1,])
```
I evaluate the results
```{r,eval=T}
par(mfrow=c(2,3))
plot(sam1[,1],c_r1[1,],ylim=c(-10,10),xlab="other",ylab="sprat difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam1[,1],c_r1[2,],ylim=c(-10,10),xlab="other",ylab="mackerel difference"
     ,main="Commercial catches")
abline(h=0,col="red")
plot(sam1[,1],c_r1[3,],ylim=c(-10,10),xlab="other",ylab="cod difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam1[,1],c_r1[4,],ylim=c(-10,10),xlab="other",ylab="sprat difference",
     main="Survey catches")
abline(h=0,col="red")
plot(sam1[,1],c_r1[5,],ylim=c(-10,10),xlab="other",ylab="mackerel difference",
     main="Survey catches")
abline(h=0,col="red")
plot(sam1[,1],c_r1[6,],ylim=c(-10,10),xlab="other",ylab="cod difference",
     main="Survey catches")
abline(h=0,col="red")
```
This implies that $\theta_1$ has to be larger than 24. For $\theta_{2:4}$:
```{r,eval=T}
par(mfrow=c(2,3))
plot(sam1[,2],c_r1[1,],ylim=c(-10,10),xlab="sprat b",ylab="sprat difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam1[,3],c_r1[2,],ylim=c(-10,10),xlab="mackerel b",ylab="mackerel difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam1[,4],c_r1[3,],ylim=c(-10,10),xlab="cod b",ylab="cod difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam1[,2],c_r1[4,],ylim=c(-10,10),xlab="sprat b",ylab="sprat difference",
     main="Survey catches")
abline(h=0,col="red")
plot(sam1[,3],c_r1[5,],ylim=c(-10,10),xlab="mackerel b",ylab="mackerel difference",
     main="Survey catches")
abline(h=0,col="red")
plot(sam1[,4],c_r1[6,],ylim=c(-10,10),xlab="cod b",ylab="cod difference",
     main="Survey catches")
abline(h=0,col="red")
```
which implies that $\theta_{2:4}$ needs to be larger than 7, 6 and 0 for sprat, mackerel and cod respectively. For $\theta_{5:7}$:
```{r,eval=T}
par(mfrow=c(2,3))
plot(sam1[,5],c_r1[1,],ylim=c(-10,10),xlab="Spin up F sprat",ylab="sprat difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam1[,6],c_r1[2,],ylim=c(-10,10),xlab="Spin up F mackerel",ylab="mackerel difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam1[,7],c_r1[3,],ylim=c(-10,10),xlab="Spin up F cod",ylab="cod difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam1[,5],c_r1[4,],ylim=c(-10,10),xlab="Spin up F sprat",ylab="sprat difference",
     main="Survey catches")
abline(h=0,col="red")
plot(sam1[,6],c_r1[5,],ylim=c(-10,10),xlab="Spin up F mackerel",ylab="mackerel difference",
     main="Survey catches")
abline(h=0,col="red")
plot(sam1[,7],c_r1[6,],ylim=c(-10,10),xlab="Spin up F cod",ylab="cod difference",
     main="Survey catches")
abline(h=0,col="red")
```
This shows that $\theta_7$, the value for cod, must be less than one. To confirm this I will run a second round.
```{r,eval=F}
sam2 <- sobol(5000,3*3 + 1,init = F)
sam2[,1] <- sam2[,1] *6 +24
sam2[,2:4] <- t(t(sam2[,2:4]) * c(13,14,20) + c(7,6,0))
sam2[,5:7] <- t(t(sam2[,5:7]) * c(1.5,1.5,1) + c(0,0,0.0))
sam2[,8:10] <- sam2[,8:10] * 1.5
round2 <- pbapply(sam2,1,hist_mat,params=params)
c_r2 <- round2 - c(obs_c[1,],obs_s[1,])
```
I evaluate the results
```{r,eval=T}
par(mfrow=c(2,3))
plot(sam2[,1],c_r2[1,],ylim=c(-10,10),xlab="other",ylab="sprat difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam2[,1],c_r2[2,],ylim=c(-10,10),xlab="other",ylab="mackerel difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam2[,1],c_r2[3,],ylim=c(-10,10),xlab="other",ylab="cod difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam2[,1],c_r2[4,],ylim=c(-10,10),xlab="other",ylab="sprat difference",
     main="Survey catches")
abline(h=0,col="red")
plot(sam2[,1],c_r2[5,],ylim=c(-10,10),xlab="other",ylab="mackerel difference",
     main="Survey catches")
abline(h=0,col="red")
plot(sam2[,1],c_r2[6,],ylim=c(-10,10),xlab="other",ylab="cod difference",
     main="Survey catches")
abline(h=0,col="red")

par(mfrow=c(2,3))
plot(sam2[,2],c_r2[1,],ylim=c(-10,10),xlab="sprat b",ylab="sprat difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam2[,3],c_r2[2,],ylim=c(-10,10),xlab="mackerel b",ylab="mackerel difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam2[,4],c_r2[3,],ylim=c(-10,10),xlab="cod b",ylab="cod difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam2[,2],c_r2[4,],ylim=c(-10,10),xlab="sprat b",ylab="sprat difference",
     main="Survey catches")
abline(h=0,col="red")
plot(sam2[,3],c_r2[5,],ylim=c(-10,10),xlab="mackerel b",ylab="mackerel difference",
     main="Survey catches")
abline(h=0,col="red")
plot(sam2[,4],c_r2[6,],ylim=c(-10,10),xlab="cod b",ylab="cod difference",
     main="Survey catches")
abline(h=0,col="red")

par(mfrow=c(2,3))
plot(sam2[,5],c_r2[1,],ylim=c(-10,10),xlab="Spin up F sprat",ylab="sprat difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam2[,6],c_r2[2,],ylim=c(-10,10),xlab="Spin up F mackerel",ylab="mackerel difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam2[,7],c_r2[3,],ylim=c(-10,10),xlab="Spin up F cod",ylab="cod difference",
     main="Commercial catches")
abline(h=0,col="red")
plot(sam2[,5],c_r2[4,],ylim=c(-10,10),xlab="Spin up F sprat",ylab="sprat difference",
     main="Survey catches")
abline(h=0,col="red")
plot(sam2[,6],c_r2[5,],ylim=c(-10,10),xlab="Spin up F mackerel",ylab="mackerel difference",
     main="Survey catches")
abline(h=0,col="red")
plot(sam2[,7],c_r2[6,],ylim=c(-10,10),xlab="Spin up F cod",ylab="cod difference",
     main="Survey catches")
abline(h=0,col="red")
```
I am happy that I have narrowed down the space and are now happy to setup the MCMC algorithm.

# Markov Chain Monte Carlo

I will first setup the moves. I first setup the likelihood of the full model and the MDA-MCMC algorithm
```{r,eval=F}
likelihood <- function(obs_c,obs_s,params,model_run,sigma_c,sigma_s,Fs,sur_ef,time=20){
  fish_rate <- cbind(Fs,sur_ef)
  colnames(fish_rate) <- NULL
  catches <- get_CPG(params,model_run,effort=fish_rate)
  com_catches_t <- apply(catches[,1:3,],c(2,3),sum)
  sur_catches <-  com_catches <- matrix(0,time,3)
  for (i in 1:time){
    com_catches[i,] <- log(rowSums(com_catches_t[,10 * (i-1) + (1:10)])) - log(1e6)
    sur_catches[i,] <- log(rowSums(catches[,4,10 * (i-1) + (1:10)])) - log(1e6)
  }
  return(rbind(dnorm(t(obs_c),t(com_catches),sigma_c,log=T),
               dnorm(t(obs_s),t(sur_catches),sigma_s,log=T)))
}

get_likelihood<- function(theta,Fs,params,sur_ef,sigma_c,sigma_s,obs_c=obs_c,obs_s=obs_s){
  params@other <- exp(theta[1])
  params@recruit_params[[1]][2] <- exp(theta[2])
  params@recruit_params[[2]][2] <- exp(theta[3])
  params@recruit_params[[3]][2] <- exp(theta[4]); 
  tmp <- theta[5:7]
  model_run <- run_LeMans(params, years=100, effort=t(matrix(c(tmp,0),4,100)))
  fish_rate <- as.matrix(cbind(Fs,sur_ef))
  colnames(fish_rate) <- NULL
  model_run_act <- run_LeMans(params, years=20, effort=fish_rate,
                              N0=model_run@N[,,dim(model_run@N)[3]])
  #browser()
  like<- likelihood(obs_c,obs_s,params,model_run_act,sigma_c,sigma_s,Fs,sur_ef)
  return(list(like=like,model_run=model_run))
}

mda_mcmc <- function(params,model_run,theta,Fs,sur_ef,prop_sdR,prop_sdF,
                     sigma_c,sigma_s,obs_c,obs_s,like){
  prop_ <- parSapply(cl,1:3,mda_mcmc1,params=params,theta=theta,Fs=Fs,
                     sur_ef=sur_ef,prop_sdR=prop_sdR,prop_sdF=prop_sdF,sigma_c=sigma_c,
                     sigma_s=sigma_s,obs_c=obs_c,obs_s=obs_s,like=like)
  thetar <- matrix(rep(c(theta[1],prop_[2,],prop_[3,]),4),4,byrow=T)
  for (i in 2:4){
    thetar[i,c(i,i + 3)] <- c(prop_[c(4,5),i-1])
  }
  like_ <- parRapply(cl,thetar,get_likelihood,Fs=Fs,params=params,sur_ef=sur_ef,
                     sigma_c=sigma_c,sigma_s=sigma_s,obs_c=obs_c,obs_s=obs_s)
  tmp <- lapply(like_[-1],function(x,y){rowSums(x[[1]]-y)},y=like_[[1]][[1]])
  tmp1 <- matrix(unlist(tmp),nrow=3,byrow = T)
  tmp <- c(tmp1[1,1] + tmp1[1,4],tmp1[2,2] + tmp1[2,5],tmp1[3,3] + tmp1[3,6])
  tmp1 <- ifelse(tmp > 0,1,exp(tmp))
  acc_rej<- sum(like_[[1]][[1]]) - sum(like) 
  - sum(log(ifelse(prop_[6,]==1,prop_[1,],1-prop_[1,]))) 
  + sum(log(ifelse(prop_[6,]==1,tmp1,1-tmp1)))
  if (log(runif(1)) < acc_rej){
    print("accept")
    return(list(thetar[1,],like_[[1]][[2]],like_[[1]][[1]]))
  }
  return(list(theta,model_run,like))
}

mda_mcmc1 <- function(i,params,theta,Fs,sur_ef,prop_sdR,prop_sdF,sigma_c,sigma_s,
                      obs_c,obs_s,like){
  theta_ <- rnorm(1,theta[i+1],prop_sdR[i])
  spin_upF_ <- abs(rnorm(1,theta[i+4],prop_sdF[i]))
  params@recruit_params[[1]][2] <- exp(theta[2])
  params@recruit_params[[2]][2] <- exp(theta[3])
  params@recruit_params[[3]][2] <- exp(theta[4]);  
  params@recruit_params[[i]][2] <- exp(theta_)
  params@other <- exp(theta[1])
  tmp <- theta[5:7]
  tmp[i] <- spin_upF_
  model_run <- run_LeMans(params, years=100, effort=t(matrix(c(tmp,0),4,100)))
  fish_rate <- as.matrix(cbind(Fs,sur_ef))
  colnames(fish_rate) <- NULL
  model_run_act <- run_LeMans(params, years=20, effort=fish_rate,
                              N0=model_run@N[,,dim(model_run@N)[3]])
  like_ <- sum(likelihood(obs_c,obs_s,params,model_run_act,
                          sigma_c,sigma_s,Fs,sur_ef)[c(i,i+3),])
  alpha <- exp(min(0,like_  - sum(like[c(i,i+3),])))
  if (runif(1) <= alpha){
    return(c(alpha,theta_,spin_upF_,theta[i+1],theta[i+4],1))
  }
  return(c(alpha,theta[i+1],theta[i+4],theta_,spin_upF_,0))
}

```
In this case the likelihood was split into 3:
\begin{equation*}
\prod{}_{t=1}^{20}N(\ln(w_{i,t})|\ln(c(M_{i,t}(\theta,\phi_{1:20,1:3}))), \sigma_{c,i}^2)N(\ln(z_{i,t})|\ln(b(M_{i,t}(\theta,\phi_{1:20,1:3}))), \sigma_{s,i}^2),
\end{equation*}
with $i=1$, $2$ and $3$, and the 3 sets of parameters $\{\theta_2,\theta_5\}$, $\{\theta_3,\theta_6\}$ and $\{\theta_4,\theta_7\}$ respectively. I setup the @caulderhead move:
```{r,eval=F}
calderhead_move <- function(params,model_run,theta,Fs,sur_ef,cd_calder_move,
                            sigma_c,sigma_s,obs_c,obs_s,like, M=10){
  theta1 <- as.numeric(theta + t(cd_calder_move) %*% rnorm(7)   )
  theta_ <- abs(cbind(theta,theta1 + t(cd_calder_move) %*% 
                        matrix(rnorm(7 * M),nrow=7,ncol=M) ))
  theta_[1,]<- ifelse(theta_[1,] > 30, 60 - theta_[1,],theta_[1,])
  like_ <- parCapply(cl,theta_,get_likelihood,Fs=Fs,params=params,
                     sur_ef=sur_ef,sigma_c=sigma_c,sigma_s=sigma_s,
                     obs_c=obs_c,obs_s=obs_s)
  tmp <- unlist(lapply(like_,function(x){sum(x[[1]])}))
  number <- sample(M+1,1,prob=exp(tmp - max(tmp)))
  print(number)
  return(list(as.numeric(theta_[,number]),like_[[number]][[2]],like_[[number]][[1]]))
}
```
The PDA-MCMC algorithm is setup as follows. First I setup the likelihood and then run through the PDA-MCMC algorithm. In the PDA-MCMC algorithm I only change a few fishing rates at a time, controlled by the `change` input. 
```{r,eval=F}
log_like <- function(model_run,i,obs_c,obs_s,sigma_c,sigma_s,fishing_rate){
  catches <- get_CPG(params,model_run,effort=fishing_rate)
  com_catches_t <- apply(catches[,1:3,],c(2,3),sum)
  com_catches <- log(rowSums(com_catches_t[,(1:10)])) - log(1e6)
  sur_catches <- log(rowSums(catches[,4,(1:10)])) - log(1e6)
  return(sum(dnorm(obs_c[i,],com_catches,sigma_c,log=T) 
             + dnorm(obs_s[i,],sur_catches,sigma_s,log=T)))
}

pda_mcmc <- function(obs_c,obs_s,theta,params,model_run,sigma_c,
                     sigma_s,Fs,sur_ef,llike,sigmasFF,change=1:20){
  params@other <- exp(theta[1])
  params@recruit_params[[1]][2] <- exp(theta[2])
  params@recruit_params[[2]][2] <- exp(theta[3])
  params@recruit_params[[3]][2] <- exp(theta[4])
  tmp <- theta[5:7]
  fish_rate <- cbind(Fs,sur_ef)
  colnames(fish_rate) <- NULL
  Q_0_ <-Q_0 <- M_0 <- model_run
  fish_rate_ <- fish_rate
  alpha <- rep(0)
  alpha1 <- rep(0)
  llike_ <- 0
  hast <- 0
    for (i in 1:20){
      if (any(change==i)){
        tmp <- rnorm(3,log(fish_rate[i,1:3]),sigmasFF)
        if (all(exp(tmp) < 2)){
          M_0_ <- run_LeMans(params, years=1, 
                             effort=t(c(exp(tmp),fish_rate[i,4])),
                             N0=M_0@N[,,dim(M_0@N)[3]])
          ll_p <- log_like(M_0_,i,obs_c,obs_s,sigma_c,sigma_s,
                           t(c(exp(tmp),fish_rate[i,4])))
          Q_0_ <- run_LeMans(params, years=1, effort=t(c(exp(tmp),fish_rate[i,4])),
                             N0=Q_0@N[,,dim(Q_0@N)[3]])
          alpha1b <-  log_like(Q_0_,i,obs_c,obs_s,sigma_c,sigma_s,
                               t(c(exp(tmp),fish_rate[i,4])))
        }
        else{
          ll_p <- -Inf
          alpha1 <- Inf
          alpha1b <- -Inf
        }
        M_0 <- run_LeMans(params, years=1, effort=t(fish_rate[i,]),
                          N0=M_0@N[,,dim(M_0@N)[3]])
        Q_0 <- run_LeMans(params, years=1, effort=t(fish_rate[i,]),
                          N0=Q_0@N[,,dim(Q_0@N)[3]])
        ll_r <- log_like(M_0,i,obs_c,obs_s,sigma_c,sigma_s,t(fish_rate[i,]))
        alpha1a <- log_like(Q_0,i,obs_c,obs_s,sigma_c,sigma_s,t(fish_rate[i,]))
        alpha1 <- exp(alpha1a - alpha1b)
        alpha <- min(1,exp(ll_p - ll_r))
        
        if( runif(1) <= alpha){
          llike_ <- llike_ + ll_p
          M_0 <- M_0_
          fish_rate_[i,] <- c(exp(tmp),fish_rate[i,4])
          hast <- hast + log(min(alpha1,1)) - log(alpha)
        }
        else{
          llike_ <- llike_ + ll_r
          fish_rate_[i,] <- fish_rate[i,]
          hast <- hast + log(1 - min(1/alpha1,1)) - log(1 - alpha)
        }
      }else{
        M_0 <- run_LeMans(params, years=1, effort=t(fish_rate[i,]),
                          N0=M_0@N[,,dim(M_0@N)[3]])
        Q_0 <- run_LeMans(params, years=1, effort=t(fish_rate[i,]),
                          N0=Q_0@N[,,dim(Q_0@N)[3]])
        llike_ <- llike_ +  log_like(M_0,i,obs_c,obs_s,sigma_c,
                                     sigma_s,t(fish_rate[i,]))
      }
    }
  if(log(runif(1)) < (llike_ - llike + hast)){
    print("accept fish_rate")
    return(list(llike=llike_,fish_rate=fish_rate_[,1:3]))
  }
  else{
    return(list(llike=llike,fish_rate=fish_rate[,1:3]))
  }
}
```
I included `print` statements in the functions so that I could monitor the efficientcy of the algorithm. The final updates were the Gibbs steps for the variance parameters.
```{r,eval=F}
gibbs_move <- function(params,model_run,theta,Fs,sur_ef){
  params@other <- exp(theta[1])
  params@recruit_params[[1]][2] <- exp(theta[2])
  params@recruit_params[[2]][2] <- exp(theta[3])
  params@recruit_params[[3]][2] <- exp(theta[4])
  tmp <- theta[5:7]
  fish_rate <- as.matrix(cbind(Fs,sur_ef))
  colnames(fish_rate) <- NULL
  model_run_act <- run_LeMans(params, years=20, effort=fish_rate,
                              N0=model_run@N[,,dim(model_run@N)[3]])
  catches <- get_CPG(params,model_run_act,effort=fish_rate)
  com_catches_t <- apply(catches[,1:3,],c(2,3),sum)
  sur_catches <-  com_catches <- matrix(0,20,3)
  for (i in 1:20){
    com_catches[i,] <- log(rowSums(com_catches_t[,10 * (i-1) + (1:10)])) 
    - log(1e6)
    sur_catches[i,] <- log(rowSums(catches[,4,10 * (i-1) + (1:10)])) 
    - log(1e6)
  }
  sigma_c <- sqrt(1/rgamma(3,shape=10 + 0.1,rate=0.1+
    rowSums((t(obs_c)-t(com_catches))^2)/2))
  sigma_s <- sqrt(1/ rgamma(3,shape=10 +0.1,rate=0.1 +
    rowSums((t(obs_s)-t(sur_catches))^2)/2))
  return(list(sigma_c,sigma_s,rbind(dnorm(t(obs_c),t(com_catches),
                                          sigma_c,log=T),dnorm(t(obs_s),
                                          t(sur_catches),sigma_s,log=T))))
}
```
The Caulderhead MCMC algorithm and the MDA-MCMC were run using parallel computing. The PDA-MCMC algorithm was not. LeMans is more computational efficient that mizer and therefore there was no computational saving by doing this algorithm in parallel.

For a wider discussion about tuning MCMC see @Bedard2019.

I chose initial parameters to start the MCMC. I got these from the history matching.
```{r,eval=F}
theta <- c(24, 10,6,0, 0.5,0.5,0.5)
Fs <- matrix(runif(20 *3 ,0,1.5),20,3)
sigma_s  <- c(0.5,0.5,0.5)
sigma_c <- c(0.5,0.5,0.5)
```
I set the tuning parameters for the MCMC algorithm, tuned from a pilot run of the MCMC. For the MDA-MCMC algorithm I set the stardard deviation of the recruitment parameters to be 0.2 and the `spinupF` to be 0.1. In this algorithm one can be bold with the moves of the parameters. The standard deviation on the natural log scale for the fish rates is 0.2. I tuned this by visually checking that the traceplots were sufficiently uncorrelated.
```{r,eval=F}
prop_sdR <- c(0.2,0.2,0.2)
prop_sdF <- c(0.1,0.1,0.1)
sigmasFF <- 0.2
```
```{r,echo=F}
load("final_lemans.Rdata")
```
I also set the `cd_calder_move` move covariance matrix. This was also tuned by visually checking that the traceplots were sufficiently uncorrelated. I set it to be
```{r}
cd_calder_move
```
`cd_calder_move` is the transpose of the Cholesky decomposition of the covariance matrix used when proposing new `theta` values. I setup variables for saving the parameters.
```{r,eval=F}
library(abind)
thetas <- theta
Fss <- array(Fs,dim=c(20,3,1))
sigmas_s <- sigma_s
sigmas_c <- sigma_c
```
and then calculate the likelihood and model after the spinup period for the current parameter settings.
```{r,eval=F}
tmp <- get_likelihood(theta,Fs,params,sur_ef,sigma_c,sigma_s,obs_c=obs_c,obs_s=obs_s)
like <-  tmp[[1]]; model_run <- tmp[[2]]
```
To get the fishing rates in the ballpark and speeding up the burn-in period of the algorithm I run a version of the PDA-MCMC algorithm, that does not deal with the Hasting ratio.
```{r,eval=F}
pda_mcmc_spin_up <- function(obs_c,obs_s,theta,params,model_run,sigma_c,
                             sigma_s,Fs,sur_ef,llike,sigmasFF,change=1:20){
    params@other <- exp(theta[1])
    params@recruit_params[[1]][2] <- exp(theta[2])
    params@recruit_params[[2]][2] <- exp(theta[3])
    params@recruit_params[[3]][2] <- exp(theta[4])
    tmp <- theta[5:7]
    fish_rate <- cbind(Fs,sur_ef)
    colnames(fish_rate) <- NULL
    Q_0_ <-Q_0 <- M_0 <- model_run
    fish_rate_ <- fish_rate
    alpha <- rep(0,20)
    alpha1 <- rep(0,20)
    llike_ <- 0
    hast <- 0
    for (i in 1:20){
      if (any(change==i)){
        tmp <- rnorm(3,log(fish_rate[i,1:3]),sigmasFF)
        if (all(exp(tmp) < 2)){
          M_0_ <- run_LeMans(params, years=1, effort=t(c(exp(tmp),fish_rate[i,4])),
                             N0=M_0@N[,,dim(M_0@N)[3]])
          ll_p <- log_like(M_0_,i,obs_c,obs_s,sigma_c,sigma_s,
                           t(c(exp(tmp),fish_rate_[i,4])))
        }
        else{
          ll_p <- -Inf
        }
        M_0 <- run_LeMans(params, years=1, effort=t(fish_rate[i,]),
                          N0=M_0@N[,,dim(M_0@N)[3]])
        ll_r <- log_like(M_0,i,obs_c,obs_s,sigma_c,sigma_s,
                         t(fish_rate[i,]))
        alpha <- min(1,exp(ll_p - ll_r))
        if( runif(1) <= alpha){
          llike_ <- llike_ + ll_p
          M_0 <- M_0_
          fish_rate_[i,] <- c(exp(tmp),fish_rate[i,4])
        }
        else{
          llike_ <- llike_ + ll_r
          fish_rate_[i,] <- fish_rate[i,]
        }
      }else{
        M_0 <- run_LeMans(params, years=1, effort=t(fish_rate[i,]),
                          N0=M_0@N[,,dim(M_0@N)[3]])
        llike_ <- llike_ +  log_like(M_0,i,obs_c,obs_s,sigma_c,
                                     sigma_s,t(fish_rate[i,]))
      }
    }
    if(log(runif(1)) < (llike_ - llike)){
      print(paste("accept fish_rate",change))
      return(list(llike=llike_,fish_rate=fish_rate_[,1:3]))
    }
    else{
      return(list(llike=llike,fish_rate=fish_rate[,1:3]))
    }
  }
  
  for(ii in 1:1000){
  test <- pda_mcmc_spin_up(obs_c,obs_s,theta,params,model_run,
                           sigma_c,sigma_s,Fs,sur_ef,llike=sum(like),
                           sigmasFF=sigmasFF,change=which(sample(c(0,1),
                                        20,replace=T,prob=c(0.8,0.2))==1))
  like <- test[[1]]; Fs <- test[[2]]
}
```
I then calculate the likelihood again and save it because in this code the `pda_mcmc` function, as written here, does not give the likelihood in the correct way.
```{r,eval=F}
tmp <- get_likelihood(theta,Fs,params,sur_ef,sigma_c,sigma_s,obs_c=obs_c,obs_s=obs_s)
like <-  tmp[[1]]; model_run <- tmp[[2]]
likes <- sum(like)
```
I setup the clusters and say that the MCMC will run for 20000 iterations.
```{r,eval=F}
N <- 2e4
library(parallel)
cl <- makeCluster(10)
clusterExport(cl=cl, c("get_likelihood", "run_LeMans","likelihood","get_CPG"))
```
I then run the MCMC algorithm. One iteration of the MCMC will update the parameters according to the MDA-MCMC algorithm; then it will do the PDA-MCMC algorithm 10 times, randomly chosing whether or not update $\phi_{t,1:3}$ at each $t$ with probability 0.7; then it will run the caulderheadparallel MCMC with 9 proposals and finally the variance parameters will be updated using the Gibbs steps. I collect the parameters, print the iteration number and every 1000 iterations I look at the traceplot of `theta`.
```{r,eval=F}
ii <- 1
while(ii < N){
  ## mda-mcmc
  test <- mda_mcmc(params,model_run,theta,Fs,sur_ef,prop_sdR,
                   prop_sdF,sigma_c,sigma_s,obs_c,obs_s,like)
  theta <- test[[1]]; model_run <- test[[2]]; like <- test[[3]]
  
  ## pda-mcmc
  for(kk in 1:10){
    test <- pda_mcmc(obs_c,obs_s,theta,params,model_run,sigma_c,
                     sigma_s,Fs,sur_ef,llike=sum(like),sigmasFF=sigmasFF,
                     change=which(sample(c(0,1),20,replace=T,prob=c(0.3,0.7))==1))
    like <- test[[1]]; Fs <- test[[2]]
  }
  ## calderhead
  test <- calderhead_move(params,model_run,theta,Fs,sur_ef,cd_calder_move,
                          sigma_c,sigma_s,obs_c,obs_s,like, M=9)
  theta <- test[[1]]; model_run <- test[[2]]; like <- test[[3]]
  
  ## gibbs move
  test <- gibbs_move(params,model_run,theta,Fs,sur_ef)
  sigma_c <- test[[1]]; sigma_s <- test[[2]]; like <- test[[3]]
  
  ## collecting
  thetas <- rbind(thetas,theta)
  Fss <- abind(Fss,Fs)
  sigmas_c <- rbind(sigmas_c,sigma_c)
  sigmas_s <- rbind(sigmas_s,sigma_s)
  likes <- c(likes,sum(like))
  if ((ii / 1000) == floor(ii / 1000)){
    plot.ts(thetas)
  }
  ii <- ii + 1
  print(ii)
}
```
I examine the traceplots for the fishing rates
```{r,eval=T}
plot.ts(t(Fss[1:10,1,]))
plot.ts(t(Fss[1:10,2,]))
plot.ts(t(Fss[1:10,3,]))
```
and for the static parameters
```{r,eval=T,figure.height=4}
plot.ts(thetas)
```

# Results

I discard the first 1000 iterations as burn up.
I examine the posterior distribution of the fishing rates, $\phi_{1:20,1:3}$. The solid line black line is the median poserior distribution and the dotted lines are the 10th and 90th percentiles. The grey line is the truth.
```{r}
par(mfrow=c(3,1))
par(mar=c(2,2,2,0))
par(oma=c(2.1,2.1,0,0))
tmp<-apply(Fss[,1,-(1:1000)],1,quantile,probs=c(0.1,0.5,0.9))
plot(c(1,20),c(0,2),type="n",main="Sprat",xlab="",ylab="")
lines(1:20,tmp[2,])
lines(1:20,tmp[1,],lty=2)
lines(1:20,tmp[3,],lty=2)
lines(1:20,fish_rate[,1],col="grey")
###
tmp<-apply(Fss[,2,-(1:1000)],1,quantile,probs=c(0.05,0.5,0.95))
plot(c(1,20),c(0,2),type="n",main="Mackerel",xlab="",ylab="")
lines(1:20,tmp[2,])
lines(1:20,tmp[1,],lty=2)
lines(1:20,tmp[3,],lty=2)
lines(1:20,fish_rate[,2],col="grey")
###
tmp<-apply(Fss[,3,-(1:1000)],1,quantile,probs=c(0.05,0.5,0.95))
plot(c(1,20),c(0,2),type="n",main="Cod",xlab="",ylab="")
lines(1:20,tmp[2,])
lines(1:20,tmp[1,],lty=2)
lines(1:20,tmp[3,],lty=2)
lines(1:20,fish_rate[,3],col="grey")
mtext("Year",side=1,outer = T,line=1)
mtext(expression(phi["1:20,1:3"]),side=2,outer = T,line=1)
```
I examine the posterior of the static parameters, $\theta$. $\theta_1$ is below with the truth being the grey line.
```{r}
par(mfrow=c(1,1))
plot(density(thetas[-(1:1000),1]),main="other",xlab=expression(theta[1]))
abline(v=27,col="grey")
```
The recruitment and spin up Fs are for Sprat
```{r}
par(mfrow=c(1,2))
plot(density(thetas[-(1:1000),2]),main="",xlab=expression(theta[2]))
abline(v=11,col="grey")
plot(density(thetas[-(1:1000),5]),main="",xlab=expression(theta[5]))
abline(v=0.1,col="grey")
```
for mackerel
```{r}
par(mfrow=c(1,2))
plot(density(thetas[-(1:1000),3]),main="",xlab=expression(theta[3]))
abline(v=7.5,col="grey")
plot(density(thetas[-(1:1000),6]),main="",xlab=expression(theta[6]))
abline(v=0.3,col="grey")
```
and for cod
```{r}
par(mfrow=c(1,2))
plot(density(thetas[-(1:1000),4]),main="",xlab=expression(theta[4]))
abline(v=1,col="grey")
plot(density(thetas[-(1:1000),7]),main="",xlab=expression(theta[7]))
abline(v=0.6,col="grey")
```
The variance parameters for sprat
```{r}
par(mfrow=c(1,2))
plot(density(sigmas_c[-(1:1000),1]),main="",xlab=expression(sigma["c,1"]))
abline(v=0.38,col="grey")
plot(density(sigmas_s[-(1:1000),1]),main="",xlab=expression(sigma["s,1"]))
abline(v=0.6,col="grey")
```
for mackerel
```{r}
par(mfrow=c(1,2))
plot(density(sigmas_c[-(1:1000),2]),main="",xlab=expression(sigma["c,2"]))
abline(v=0.44,col="grey")
plot(density(sigmas_s[-(1:1000),2]),main="",xlab=expression(sigma["s,2"]))
abline(v=0.7,col="grey")
```
and cod
```{r}
par(mfrow=c(1,2))
plot(density(sigmas_c[-(1:1000),3]),main="",xlab=expression(sigma["c,3"]))
abline(v=0.5,col="grey")
plot(density(sigmas_s[-(1:1000),3]),main="",xlab=expression(sigma["s,3"]))
abline(v=0.6,col="grey")
```

